---
title: "IDW model for short-term and long-term sensors"
author: "Jordan Wingenroth"
date: "May 31, 2018"
output: html_document
---

```{r knitr opts, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries}

library(tidyverse)
library(readxl)
library(lubridate)
library(maps)
library(gstat)
library(sp)

```

# geographic data setup

It seemed to make the most sense to load the pm 2.5 data from longterm sensors at the same time since the lat/long was included in the same table. The pm 2.5 dataset was a large file (10 MB) since it included all October data for the entire state, so I filtered it to only include sites within ~1 deg of the centerpoint of sampling sites in Excel prior to adding to the project.

```{r load farm and sensor geographic data}

st_sensors <- read_csv(file = "../data/aq_sensors.csv")
st_sensors <- filter(st_sensors, `sensor class` == "short_term")

long_term_pm25 <- read_csv("../data/long_term_pm25.csv")
lt_sensors <-long_term_pm25 %>%
  distinct(name, .keep_all = TRUE) %>%
  transmute(name, "sensor class" = "long_term", latitude, longitude)

farms <- read_csv(file = "../data/farm_data.csv")
hysplit <- readxl::read_xlsx(path = "../data/HYSPLIT_data.xlsx")

all(sort(hysplit$Address)==sort(farms$address))

```

```{r tidy geographic data}

#farm

farms <- left_join(farms, hysplit, by = c("address" = "Address"))

farms <- farms %>%
  select(address, Key, long, lat, `Exposure (raw, final model, normalized)`, `Exposure (smoothed, final model, normalized)`, `Rank (raw)`, `Rank (smoothed)`)

rm(list = "hysplit")

#aq_sensors

aq_sensors <- rbind(lt_sensors, st_sensors)
rm(list = c("lt_sensors", "st_sensors"))
aq_sensors <- rename(aq_sensors, lat = latitude, long = longitude)

```

# map with farm sites and aq sensors

```{r map, fig.width=8, fig.height=8}

ggplot(NULL, aes(long, lat)) +
  geom_point(data = aq_sensors) +
  geom_text(data = aq_sensors, color = "black", alpha = .3, aes(label = name), hjust = "inward", vjust = "inward", size = 3) +
  geom_point(data = farms, color = "red") +
  geom_text(data = farms, color = "red", alpha = .3, aes(label = Key), hjust = "inward", vjust = "inward", size = 3) +
  borders("state", "California") +
  coord_fixed(xlim = c(-123.5,-121.5), ylim = c(37.5, 39.5))

```

We're lacking sensors to the northwest of the sites, but we have a good number to the southeast (Vacaville, Vallejo, Berkeley, SF, Oakland).

ARB 14 & 38 Riebli sites have incorrect GPS data???

#fix Riebli data

```{r fix}

#values from Google Maps
aq_sensors[grep("Riebli", aq_sensors$name),]$lat <- 38.500
aq_sensors[grep("Riebli", aq_sensors$name),]$long <- -122.741

```

```{r remap}

ggplot(NULL, aes(long, lat)) +
  geom_point(data = aq_sensors) +
  geom_text(data = aq_sensors, color = "black", alpha = .3, aes(label = name), hjust = "inward", vjust = "inward", size = 3) +
  geom_point(data = farms, color = "red") +
  geom_text(data = farms, color = "red", alpha = .3, aes(label = Key), hjust = "inward", vjust = "inward", size = 3) +
  borders("state", "California") +
  coord_fixed(xlim = c(-123.5,-121.5), ylim = c(37.5, 39.5))

```


# PM 2.5 data setup

```{r load short term pm 2.5 data}

st_files <- list.files("../data/short_term_pm25", full.names = TRUE)
short_term_pm25 <- lapply(st_files, function(x) read_xlsx(x))
names(short_term_pm25) <- str_sub(st_files, 42, -20)
for (i in 1:length(short_term_pm25)) short_term_pm25[[i]] <- mutate(short_term_pm25[[i]], name = names(short_term_pm25[i]))
short_term_pm25 <- do.call(rbind, short_term_pm25)
short_term_pm25 <- select(short_term_pm25, name, everything())

sort(unique(short_term_pm25$name))==sort(aq_sensors[aq_sensors$`sensor class`=="short_term",]$name)

```

Some of the short term sensors have different, but easily matchable names.

```{r tidy pm 2.5 data}

#short-term times

short_term_pm25 <- short_term_pm25 %>%
  mutate(datetime = `Date/Time/` + 3600*hour(PST))


#long-term times

long_term_pm25 <- long_term_pm25 %>%
  mutate(date = as.POSIXct(long_term_pm25$date, format = "%m/%d/%Y"), datetime = date + 3600*start_hour)

#match sites to aq sensors by key

aq_sensors$key <- str_sub(aq_sensors$name, end = 10)
long_term_pm25$key <- str_sub(long_term_pm25$name, end = 10)
short_term_pm25$key <- str_sub(short_term_pm25$name, end = 10)

sum(sort(unique(aq_sensors$key))==sort(c(unique(long_term_pm25$key), unique(short_term_pm25$key))))

#pull out and match up essential columns for purpose of rowbinding and joining

t1<-short_term_pm25 %>%
  select(key, name, datetime, value = ConcHr)

t2<-long_term_pm25 %>%
  select(key, name, datetime, value)

#rowbind and join to gps data

pm25 <- rbind(t1, t2)

pm25 <- pm25 %>% 
  left_join(aq_sensors, "key") %>%
  select(-name.y) %>%
  rename(name = name.x)

rm(list = c("aq_sensors","long_term_pm25","short_term_pm25","t1","t2"))

#Filter out negative values and zero values as they could be errors and should be bounded by small nummbers (safe)
#Also filter out some Vacaville values that appear to be errors (>900 ug/m3)

pm25 <- pm25 %>%
  filter(value>=0, value<900)

```

# convert lat/long to kilometers for all data

Data from http://www.csgnetwork.com/degreelenllavcalc.html

```{r}

farms<- farms %>%
  mutate(northing_km = 111.005*(lat-38.5), easting_km = 87.233*(long+122.75))

pm25 <- pm25 %>%
  mutate(northing_km = 111.005*(lat-38.5), easting_km = 87.233*(long+122.75))

ggplot(NULL, aes(easting_km, northing_km)) +
  geom_point(data = distinct(pm25,long,lat,.keep_all=TRUE)) +
  geom_text(data = distinct(pm25,long,lat,.keep_all=TRUE), color = "black", alpha = .3, aes(label = name), hjust = "inward", vjust = "inward", size = 3) +
  geom_point(data = farms, color = "red") +
  geom_text(data = farms, color = "red", alpha = .3, aes(label = Key), hjust = "inward", vjust = "inward", size = 3) +
  coord_fixed(xlim=c(-110,110),ylim=c(-110,110))

```


Q for Vanessa: any idea as to where the epicenter of burned industrial facilities was? That would have some significance for our choice of GPS location.

Currently using Larkfield-Wikiup for convenience (even lat/long)

# graphs of sensor data

```{r graph}

pm25 %>%
  filter(datetime>as.Date("2017-10-1"), datetime<as.Date("2017-10-31")) %>%
  group_by(name) %>%
  ggplot(aes(datetime, value, color = name)) +
  geom_point(alpha = .2) +
  theme(legend.position = "none")

```

The only long term and short term sensors in close proximity were the Napa ones. Let's compare:

```{r napa comparison}

#Set to start on Oct 12 because that's when the short term Napa sensor was placed
pm25 %>%
  filter(datetime>as.Date("2017-10-12"), datetime<as.Date("2017-10-31"), str_detect(name, "Nap")) %>%
  group_by(name) %>%
  ggplot(aes(datetime, value, color = name)) +
  geom_smooth()

```

Reasonably similar, especially given their slight geographic distance.

# stats on October sensor data

```{r stats}

pm25 %>%
  filter(datetime>as.Date("2017-10-1"), datetime<as.Date("2017-10-31")) %>%
  group_by(name) %>%
  summarize("n of measurements" = n(), mean = mean(value), min = min(value), Q1 = quantile(value, .25), median = median(value), Q3 = quantile(value, .75), max = max(value)) %>%
  arrange(desc(mean))

```

# sensor data coverage in October

```{r heatmap}

#going to go with a range of 10/2-10/30 because sensors are running more or less dependably on that span

pm25 %>%
  filter(datetime>as.Date("2017-10-2"), datetime<as.Date("2017-10-30")) %>%
  group_by(name) %>%
  ggplot(aes(x = name, y = datetime)) +
  geom_raster(aes(fill = value)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r count over time graph}

pm25 %>%
  filter(datetime>as.Date("2017-10-2"), datetime<as.Date("2017-10-30")) %>%
  group_by(datetime) %>%
  summarise(n = n()) %>%
  ggplot(aes(datetime, n)) +
  geom_line()

```


# model function

```{r IDW model}

IDW <- function(data = pm25, start_time = "2017-10-2", end_time = "2017-10-30", xlim = c(), ylim = c()) {
  for (i in seq.POSIXt(as.POSIXct(start_time), as.POSIXct(end_time), "hour")) {
    
    timepoint <- data %>%
      filter(datetime==i)
    
    idw0(value~1, data = timepoint)
      
  }
}

```


# run model

# model analysis and plots

# get values for farms