---
title: "IDW model for short-term and long-term sensors"
author: "Jordan Wingenroth"
date: "May 31, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r knitr opts, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

```

```{r libraries}

library(tidyverse)
library(readxl)
library(lubridate)
library(maps)
library(gstat)
library(animation)
library(gganimate)
library(sp)
library(knitr)
library(raster)

```

# geographic data setup

It seemed to make the most sense to load the pm 2.5 data from longterm sensors 
at the same time since the lat/long was included in the same table. The pm 2.5 
dataset was a large file (10 MB) since it included all October data for the 
entire state, so I filtered it to only include sites within ~1 deg of the 
centerpoint of sampling sites in Excel prior to adding to the project.

```{r load farm and sensor geographic data}

st_sensors <- read_csv(file = "../data/aq_sensors.csv")
st_sensors <- filter(st_sensors, `sensor class` == "short_term")

long_term_pm25 <- read_csv("../data/long_term_pm25.csv")
lt_sensors <-long_term_pm25 %>%
  distinct(name, .keep_all = TRUE) %>%
  transmute(name, "sensor class" = "long_term", latitude, longitude)

farms <- read_csv(file = "../data/farm_data.csv")
hysplit <- readxl::read_xlsx(path = "../data/HYSPLIT_data.xlsx")

all(sort(hysplit$Address)==sort(farms$address))

```

```{r tidy geographic data}

#farm

farms <- left_join(farms, hysplit, by = c("address" = "Address"))

farms <- farms %>%
  dplyr::select(address, Key, long, lat, 
                `Exposure (raw, final model, normalized)`, 
                `Exposure (smoothed, final model, normalized)`, 
                `Rank (raw)`, `Rank (smoothed)`)

rm(list = "hysplit")

#aq_sensors

aq_sensors <- rbind(lt_sensors, st_sensors)
rm(list = c("lt_sensors", "st_sensors"))
aq_sensors <- rename(aq_sensors, lat = latitude, long = longitude)

```

# map with farm sites and aq sensors

```{r map, fig.width=8, fig.height=8}

ggplot(NULL, aes(long, lat)) +
  geom_point(data = aq_sensors) +
  geom_text(data = aq_sensors, color = "black", alpha = .3, aes(label = name), 
            hjust = "inward", vjust = "inward", size = 3) +
  geom_point(data = farms, color = "red") +
  geom_text(data = farms, color = "red", alpha = .3, aes(label = Key), 
            hjust = "inward", vjust = "inward", size = 3) +
  borders("state", "California") +
  coord_fixed(xlim = c(-123.5,-121.5), ylim = c(37.5, 39.5))

```

We're lacking sensors to the northwest of the sites, but we have a good number
to the southeast (Vacaville, Vallejo, Berkeley, SF, Oakland).

ARB 14 & 38 Riebli sites have incorrect GPS data???

#fix Riebli data

```{r fix}

#values from Google Maps
aq_sensors[grep("Riebli", aq_sensors$name),]$lat <- 38.500
aq_sensors[grep("Riebli", aq_sensors$name),]$long <- -122.741

```

```{r remap}

ggplot(NULL, aes(long, lat)) +
  geom_point(data = aq_sensors) +
  geom_text(data = aq_sensors, color = "black", alpha = .3, aes(label = name), 
            hjust = "inward", vjust = "inward", size = 3) +
  geom_point(data = farms, color = "red") +
  geom_text(data = farms, color = "red", alpha = .3, aes(label = Key), 
            hjust = "inward", vjust = "inward", size = 3) +
  borders("state", "California") +
  coord_fixed(xlim = c(-123.5,-121.5), ylim = c(37.5, 39.5))

```


# PM 2.5 data setup

```{r load short term pm 2.5 data}

st_files <- list.files("../data/short_term_pm25", full.names = TRUE)
short_term_pm25 <- lapply(st_files, function(x) read_xlsx(x))
names(short_term_pm25) <- str_sub(st_files, 42, -20)
for (i in 1:length(short_term_pm25)) {
  
  short_term_pm25[[i]] <- mutate(short_term_pm25[[i]], 
                                 name = names(short_term_pm25[i]))
  
}

short_term_pm25 <- do.call(rbind, short_term_pm25)
short_term_pm25 <- dplyr::select(short_term_pm25, name, everything())

sort(unique(short_term_pm25$name))==sort(aq_sensors[aq_sensors$`sensor class`=="short_term",]$name)

```

Some of the short term sensors have different, but easily matchable names.

```{r tidy pm 2.5 data}

#short-term times

short_term_pm25 <- short_term_pm25 %>%
  mutate(datetime = `Date/Time/` + 3600*hour(PST))


#long-term times

long_term_pm25 <- long_term_pm25 %>%
  mutate(date = as.POSIXct(long_term_pm25$date, format = "%m/%d/%Y"), datetime = date + 3600*start_hour)

#match sites to aq sensors by key

aq_sensors$key <- str_sub(aq_sensors$name, end = 10)
long_term_pm25$key <- str_sub(long_term_pm25$name, end = 10)
short_term_pm25$key <- str_sub(short_term_pm25$name, end = 10)

sum(sort(unique(aq_sensors$key))==sort(c(unique(long_term_pm25$key), unique(short_term_pm25$key))))

#pull out and match up essential columns for purpose of rowbinding and joining

t1<-short_term_pm25 %>%
  dplyr::select(key, name, datetime, value = ConcHr)

t2<-long_term_pm25 %>%
  dplyr::select(key, name, datetime, value)

#rowbind and join to gps data

pm25 <- rbind(t1, t2)

pm25 <- pm25 %>% 
  left_join(aq_sensors, "key") %>%
  dplyr::select(-name.y) %>%
  rename(name = name.x)

rm(list = c("aq_sensors","long_term_pm25","short_term_pm25","t1","t2"))

#Filter out negative values and zero values as they could be errors and should be bounded by small nummbers (safe)
#Also filter out some Vacaville values that appear to be errors (>900 ug/m3)

pm25 <- pm25 %>%
  filter(value>=0, value<900)

```

# convert lat/long to kilometers for all data

Data from http://www.csgnetwork.com/degreelenllavcalc.html

```{r northing easting}

farms<- farms %>%
  mutate(northing_km = 111.005*(lat-38.5), easting_km = 87.233*(long+122.75))

pm25 <- pm25 %>%
  mutate(northing_km = 111.005*(lat-38.5), easting_km = 87.233*(long+122.75))

pm25 <- pm25 %>%
  filter(key!="ARB 33 San", !str_detect(key, " Rin"), !str_detect(key, "Sebast"))

ggplot(NULL, aes(easting_km, northing_km)) +
  geom_point(data = distinct(pm25,long,lat,.keep_all=TRUE)) +
  geom_text(data = distinct(pm25,long,lat,.keep_all=TRUE), color = "black", alpha = .3, aes(label = name), hjust = "inward", vjust = "inward", size = 3) +
  geom_point(data = farms, color = "red") +
  geom_text(data = farms, color = "red", alpha = .3, aes(label = Key), hjust = "inward", vjust = "inward", size = 3) +
  coord_fixed(xlim=c(-110,110),ylim=c(-110,110))

```


Q for Vanessa: any idea as to where the epicenter of burned industrial facilities was? That would have some significance for our choice of GPS location.

Currently using Larkfield-Wikiup for convenience (quarter-degree lat/long)

# graphs of sensor data

```{r graph}

pm25 %>%
  filter(datetime>as.Date("2017-10-1"), datetime<as.Date("2017-10-31")) %>%
  group_by(name) %>%
  ggplot(aes(datetime, value, color = name)) +
  geom_point(alpha = .2) +
  theme(legend.position = "none")

```

From this, I've arrived at the conclusion that Oct 8 - Oct 20 is a good span for the model run in order to capture the most serious part of the fire.

The only long term and short term sensors in close proximity were the Napa ones. Let's compare:

```{r napa comparison}

#Set to start on Oct 12 because that's when the short term Napa sensor was placed
pm25 %>%
  filter(datetime>as.Date("2017-10-12"), datetime<as.Date("2017-10-31"), str_detect(name, "Nap")) %>%
  group_by(name) %>%
  ggplot(aes(datetime, value, color = name)) +
  geom_smooth() +
  geom_point()

```

Reasonably similar, especially given their slight geographic distance.

# stats on October sensor data

```{r stats}

pm25 %>%
  filter(datetime>as.Date("2017-10-1"), datetime<as.Date("2017-10-31")) %>%
  group_by(name) %>%
  summarize("n of measurements" = n(), mean = mean(value), min = min(value), Q1 = quantile(value, .25), median = median(value), Q3 = quantile(value, .75), max = max(value)) %>%
  arrange(desc(mean))

```

# sensor data coverage in October

```{r heatmap}

#going to go with a range of 10/2-10/30 because sensors are running more or less dependably on that span

pm25 %>%
  filter(datetime>as.Date("2017-10-2"), datetime<as.Date("2017-10-30")) %>%
  group_by(name) %>%
  ggplot(aes(x = name, y = datetime)) +
  geom_raster(aes(fill = value)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r count over time graph}

pm25 %>%
  filter(datetime>as.Date("2017-10-2"), datetime<as.Date("2017-10-30")) %>%
  group_by(datetime) %>%
  summarise(n = n()) %>%
  ggplot(aes(datetime, n)) +
  geom_line()

```


# model function

```{r IDW model}

myIDW <- function(data = pm25, start_time = "2017-10-08", end_time = "2017-10-20",
                xmin = -110, xmax = 110, xinc = 1, ymin = -110, ymax = 110, yinc = 1,
                idp = 2, nmax = Inf, mdist = Inf) {
  
  stack <- list()
  
  j <- 0
  
  grid <- expand.grid(x = seq(xmin, xmax, xinc),
                      y = seq(ymin, ymax, yinc))
  
  coordinates(grid) <- ~x+y
  
  gridded(grid) <- TRUE
  
  for (i in seq.POSIXt(as.POSIXct(start_time), as.POSIXct(end_time), "hour")) {
    
    timepoint <- data %>%
      filter(datetime==i)
    
    coordinates(timepoint) <- ~easting_km+northing_km
    
    j <- j+1
    
    stack[[j]] <- as_tibble(idw(value~1, locations = timepoint, newdata = grid, idp = idp, nmax = nmax, maxdist = mdist, debug.level = 0))
    stack[[j]][4] <- as.POSIXct(start_time) + 3600*(j-1)
    colnames(stack[[j]]) <- c("raster_x", "raster_y", "value", "datetime")
    
  }
  
  return(stack)
  
}

```


# With all sensors

```{r all sensors model run}

invisible(krige <- myIDW(xmin = -17, xmax = 12, xinc = 1, ymin = -32, ymax = 10, yinc = 1, idp = 2, nmax = Inf))

krige <- do.call(rbind, krige)

cumul <- krige %>%
  group_by(raster_x, raster_y) %>%
  summarize(total = sum(value), avg = mean(value), max = max(value))

```

```{r all sensors plot, fig.height= 5, fig.width=4}

cumul %>%
  mutate(lat = raster_y/111.005+38.5, long = raster_x/87.233-122.75) %>%
  ggplot(aes(y = lat, x = long)) +
  scale_fill_viridis_c() +
  geom_raster(aes(fill = total/max(total))) +
  geom_contour(color = "gray", aes(z = total)) +
  geom_point(data = farms, aes(farms$long, farms$lat), color = "white") +
  geom_point(data = distinct(pm25,lat,long,.keep_all=TRUE), aes(long, lat), color = "red") +
  coord_fixed(xlim=c(-122.95, -122.6),ylim=c(38.2, 38.6)) +
  labs(x = "long", y = "lat") +
  guides(fill = guide_colorbar(title = "rel. exp."))

ggsave("plot.png")

```

```{r all sensors farm data}

farms <- farms %>%
  mutate(raster_x = round(easting_km), raster_y = round(northing_km)) %>%
  left_join(cumul, c("raster_x", "raster_y"))

```

# Analyzing farm data

```{r tidy and print farm data}

farms %>%
  mutate(aq_norm = total/max(total)) %>%
  dplyr::select(Key, address, lat, long, 
         HYSPLIT_raw = `Exposure (raw, final model, normalized)`, 
         all_sensors = aq_norm) %>%
  mutate(HYSPLIT_raw = (HYSPLIT_raw-min(HYSPLIT_raw))/(max(HYSPLIT_raw)-min(HYSPLIT_raw)),
         all_sensors = (all_sensors-min(all_sensors))/(max(all_sensors)-min(all_sensors))) %>%
  ggplot(aes(x = HYSPLIT_raw, y = all_sensors, label = Key)) +
  geom_text(size = 2)

farmdata3_23 <- farms %>%
  rename(HYSPLIT = `Exposure (raw, final model, normalized)`) %>%
  dplyr::select(Key, avg, max) %>%
  filter(Key %in% c("SRO10", "SRO02", "ROH02", "PTL04", "SRO09", "LKW01"))

write_csv(farmdata3_23, "farmdata3_23.csv")



  
```



